---
title: "Untitled"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Revisit the PDB

> Q1: Download a CSV file from the PDB site (accessible from "Analyze" -> "PDB Statistics" > "by Experimental Method and Molecular Type". Move this CSV file into your RStudio project and determine the percentage of structures solved by X-Ray and Electron Microscopy. 

Downloaded a CSV file from: https://www.rcsb.org/stats/summary on 2019-05-08

```{r}
db <- read.csv("Data Export Summary.csv",row.names=1)
(db$Total/sum(db$Total))*100
```

Also can you determine what proportion of structures are protein?

```{r}
(sum(db$Proteins)/sum(db$Total))*100
```

## Section 3. Using Bio3D

Let's read the PDB fiLe we downloaded into R to see how things live in this place

```{r}
library(bio3d)
pdb <- read.pdb("1hsg.pdb")
pdb
```

```{r}
aa321(pdb$seqres)
```

```{r}
pdb$atom
```

### Atom selection examples

Let's select residu 10

```{r}
inds <- atom.select(pdb, resno = 1)
inds$atom
pdb$atom[inds$atom,]
```

### Make a protein only and ligand only PDB file

Here we need to select our protein with `atom.select()` then use `write.pdb()`

```{r}
p_pdb <- atom.select(pdb,"protein",value = TRUE)
write.pdb(p_pdb,"1hsg_protein.pdb")
l_pdb <- atom.select(pdb,"ligand",value = TRUE)
write.pdb(l_pdb, "1hsg_ligand.pdb")
```

```{r}
# Load the package
library("bio3d.view")
# view the 3D structure
view(pdb, "overview", col="sse")
```

## Section 6. Working with multiple PDB files

### Aligning multiple structures

```{r}
# Download some example PDB files
ids <- c("1TND_B","1AGR_A","1TAG_A","1GG2_A","1KJY_A","4G5Q_A")
files <- get.pdb(ids, split = TRUE)
```

```{r}
# Extract and align the chains we are interested in
pdbs <- pdbaln(files, fit = TRUE)
```

```{r}
# Access the first 5 rows, and 8 columns
pdbs$ali[1:5, 1:8]
# Associated residues numbers
pdbs$resno[1:5, 1:8]
```

```{r}
# Perform PCA
pc.xray <- pca(pdbs)
# Plot our results summary (PCA score plot and scree-plot)
plot(pc.xray)

```

