---
title: 'class 9: Unsupervised learning'
author: "Xiaohui Lyu"
date: "2019/5/1"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preparing the data

I download the SCV format data file from the class website and placed it in my project directory

```{r}
wisc.df <- read.csv("WisconsinCancer.csv")
head(wisc.df)
```

> Q3. How many of the observations have a malignant diagnosis? 

```{r}
table(wisc.df$diagnosis)
length(wisc.df$diagnosis[wisc.df$diagnosis=="M"])
```

> Q4. How many variables/features in the data are suffixed with _mean?

```{r}
colnames(wisc.df)
grep("_mean",colnames(wisc.df))
length(grep("_mean",colnames(wisc.df)))
```

Select cols 3:32 and store as a matrix

```{r}
wisc.data <- as.matrix(wisc.df[,3:32])
wisc.data
row.names(wisc.data) <- wisc.df$id
diagnosis <- wisc.df$diagnosis
```



## Principal Component Analysis (PCA)

Do I need to scale the data befor PCA? Check the col means and standard deviations.

```{r}
round(apply(wisc.data,2,sd))
round(colMeans(wisc.data))
```

Looks like I need to scale this data before PCA or just set the `scale=TRUE` argument to `prcomp()`

```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
summary(wisc.pr)
```

```{r}
biplot(wisc.pr)
# Scatter plot observations by components 1 and 2
plot( x=wisc.pr$x[,1],y=wisc.pr$x[,2] , col = diagnosis, 
     xlab = "PC1", ylab = "PC2")
```

```{r}
## ggplot based graph
#install.packages("factoextra")
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

```{r}
sort(abs(wisc.pr$rotation[,1]))
```



## Hierarchical clustering

```{r}
# Scale the wisc.data data: data.scaled
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist)
```

```{r}
plot(wisc.hclust)
abline(h=19, col="red",lty=2)
```

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust,k=4)
table(wisc.hclust.clusters,diagnosis)
```

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust,k=10)
table(wisc.hclust.clusters,diagnosis)
```



## Combining method

I am going to start with the PCs that capture 90% of the original variance in the dataset (i.e. PC1 to PC7)

```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:7]), method = "ward.D2")
plot(wisc.pr.hclust)
```

```{r}
grps <- cutree(wisc.pr.hclust,k=2)
table(grps, diagnosis)
```

```{r}
plot(wisc.pr$x[,1],wisc.pr$x[,2],col=grps)
```

```{r}
new <- read.csv("new_samples.csv")
npc <- predict(wisc.pr, newdata=new)
npc
plot(wisc.pr$x[,1],wisc.pr$x[,2],col=grps)
points(npc[,1],npc[,2],col="blue",pch=16,cex=3)
text(npc[,1],npc[,2],c(1,2),col="white")
```
