---
title: "Class 14: Transcriptomics and the analysis of RNA-Seq data"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 2. Import our countData and colData into R

```{r}
counts <- read.csv("data/airway_scaledcounts.csv", stringsAsFactors = FALSE)
metadata <- read.csv("data/airway_metadata.csv", stringsAsFactors = FALSE)
head(counts)
head(metadata)
```

> Q1. How many genes are in this dataset?

```{r}
nrow(counts)
```

> Q2. How many 'control' cell lines do we have?

```{r}
table(metadata$dex)
```

Let's check if the cols of `counts` is the same as the id column in `metadata`

```{r}
all (colnames(counts)[-1] == metadata$id)
```

## 3. Try differential gene expression

Let's access just their control counts data so we can calculate an average counts value for each gene in these control samples

```{r}
control <- metadata[metadata$dex == "control",]
```

Now we want to get the cols in counts with these `control$id` colnames

```{r}
control.mean <- rowSums( counts[ ,control$id] )/ nrow(control)
names(control.mean) <- counts$ensgene
```

Now do the same for the `treated` samples.

```{r}
treated <- metadata[metadata$dex == "treated",]
treated.mean <- rowSums(counts[,treated$id])/ nrow(treated)
names(treated.mean) <- counts$ensgene
head(treated.mean)
```

Let's put these together into a new data.frame called `meancounts`

```{r}
meancounts <- data.frame(control.mean, treated.mean)
colSums(meancounts)
```

```{r}
plot(control.mean, treated.mean)
```

Plot on log scale for both axis

```{r}
plot(control.mean, treated.mean, log="xy")
```

```{r}
meancounts$log2fc <- log2(meancounts$treated.mean/meancounts$control.mean)
View(meancounts)
```

```{r}
# Identify the row that contains zero
zero.vals <- which(meancounts[,1:2]==0, arr.ind = TRUE)
#Remove these rows as we can't really say much about them
to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
```

A common threshold used for calling something differentially expressed is a log2(FoldChange) of greater than 2 or less than -2. Let’s filter the dataset both ways to see how many genes are up or down-regulated.

> Q8. Using the up.ind vector above can you determine how many up regulated genes we have at the greater than 2fc level?

```{r}
up.ind <- mycounts$log2fc>2
down.ind <- mycounts$log2fc<(-2)
sum(up.ind)
sum(down.ind)
```

## 4. Adding annotation data

```{r}
anno <- read.csv("data/annotables_grch38.csv")
head(anno)
```

```{r}
mycounts.anno <- merge(mycounts, anno, by.x= 0, by.y= "ensgene")
```

## We can use bioconductor to do this too.

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

```{r}
columns(org.Hs.eg.db)
```

```{r}
mycounts$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(mycounts), # Our genenames
                     keytype="ENSEMBL",        # The format of our genenames
                     column="SYMBOL",          # The new format we want to add
                     multiVals="first")
```

```{r}
mycounts$entrez <- mapIds(org.Hs.eg.db,
                     keys=row.names(mycounts), # Our genenames
                     keytype="ENSEMBL",        # The format of our genenames
                     column="ENTREZID",          # The new format we want to add
                     multiVals="first")
```

```{r}
mycounts$uniprot <- mapIds(org.Hs.eg.db,
                     keys=row.names(mycounts),
                     column="UNIPROT",
                     keytype="ENSEMBL",
                     multiVals="first")
```

```{r}
View( mycounts[up.ind,] )
```

## 5. DESeq2 analysis

```{r}
library("DESeq2")
```

```{r}
citation("DESeq2")
```

Build our deseq dataset object need for running deseq analysis

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex, 
                              tidy=TRUE)
dds
```

```{r}
dds <- DESeq(dds)
```

Get our resluts

```{r}
res <- results(dds)
```

```{r}
summary(res)
```

```{r}
res05 <- results(dds,alpha = 0.05)
summary(res05)
```

```{r}
plot(res$log2FoldChange,-log(res$padj),
     xlab="Log2(FoldChange)",
     ylab="-Log(P-value)")
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```

```{r}
mycols <- rep("gray", nrow(res))
mycols[abs(res$log2FoldChange) > 2] <- "red"
mycols[(abs(res$log2FoldChange) > 2) & ((res$padj)< 0.01)] <- "blue"
plot(res$log2FoldChange,-log(res$padj),
     xlab="Log2(FoldChange)",
     ylab="-Log(P-value)",
     col=mycols)
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```

Another way to plot this is with the Enhanced

But first let's put proper genenames on our results (i.e. gene 'symbols')

```{r}
x <- as.data.frame(res)
x$symbol <- mapIds(org.Hs.eg.db, 
                   keys=row.names(x),
                   keytype="ENSEMBL",
                   column="SYMBOL",
                   multiVals="first")
```

```{r}
library(EnhancedVolcano)
EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')
```

