---
title: "Class 7: R functions and packages"
author: "Xiaohui Lyu"
date: "2019/4/24"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## More on function writting

First we weill revisit our function from last day
```{r}
source("http://tinyurl.com/rescale-R")
```

Test the **rescale()** function

```{r}
rescale(1:10)
```

```{r}
#rescale(c(1:10,"string"))
```

```{r}
x <- c(1:10,"string")
!is.numeric(x)
##rescale2(x)
```

## Function practice

Write a function to identify NA elements in two vectors

Start with a simple example input where I know what the answer should be.

```{r}
x <- c( 1, 2, NA, 3, NA)
y <- c(NA, 3, NA, 3, 4)
is.na(x)
is.na(y)
```

I am looking for the position where it is TRUE in both vectors.

```{r}
is.na(x) & is.na(y)
```

Take the sum to find how many

```{r}
sum(is.na(x) & is.na(y))
```

This is my working snippet of code!!

```{r}
both_na <- function(x,y){
  sum(is.na(x)& is.na(y))
}
```

```{r}
both_na(x,y)
both_na(c(NA,NA,NA),c(NA,NA,1))
both_na(c(NA,NA,NA),c(NA,NA,1,NA))
```

Check the length of inputs are equal 

```{r}
x <- c(NA,NA,NA)
y <- c(NA,NA,1,NA,NA,NA)
length(x)
length(y)
length(x) == length(y)
length(x) != length(y)
```

Add the checking into the function

```{r}
both_na2 <- function(x, y) {
  if(length(x) != length(y)) {
    stop("Input x and y should be the same length")
  }
  sum( is.na(x) & is.na(y) )
}
```

Try both_na3 with extra features

```{r}
both_na3 <- function(x, y) {
  if(length(x) != length(y)) {
    stop("Input x and y should be vectors of the same length")
  }
  na.in.both <- ( is.na(x) & is.na(y) )
  na.number <- sum(na.in.both)
  na.which <- which(na.in.both)
  message("Found ", na.number, " NA's at position(s):",
  paste(na.which, collapse=", ") )
  return( list(number=na.number, which=na.which) )
}
```

```{r}
x <- c( 1, 2, NA, 3, NA)
y <- c(NA, 3, NA, 3, 4)
both_na3(x,y)
```

## Example grade function

```{r}
grade <- function(x){
  (sum(x,na.rm = TRUE)-min(x,na.rm=TRUE))/(length(x)-1)
}
```

```{r}
grade(c(100,100,100,100,100,100,100,90))
grade(c(100,NA,90,90,90,90,97,80))
```

```{r}
students <- read.csv("student_homework.csv",row.names = 1)
ans <- apply(students,1,grade)
sort(ans,decreasing = TRUE)
```

## One last function

Find the intersection of two sets

```{r}
source("http://tinyurl.com/rescale-R")
x <- df1$IDs
y <- df2$IDs

intersect(x,y)
```

learn intersect from google and then find *%in%* in see also which is better

```{r}
x
y

x[x %in% y]
y %in% x
```

select the sentences and choose from **code -> extract function**

```{r}
gene_intersect <- function(x, y) {
  cbind(x[x %in% y],
        y[y %in% x])
}
```

```{r}
merge(df1,df2,by="IDs")
```

